ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-26 papers/GPUScA-2011/foropencl.tex
Version to ship to Matt.  Needs a couple of bib entries.

\title{Exploiting Array Syntax in Fortran for Accelerator Programming}

@@

\author{
% 1st. author
\alignauthor
Matthew J. Sottile\\
       \affaddr{Galois, Inc.}\\
       \affaddr{421 SW 6th Ave. Suite 300}\\
       \affaddr{Portland, OR 97204}\\
       \email{matt@galois.com}
% 2nd. author
\alignauthor
Craig E Rasmussen\\
       \affaddr{Los Alamos National Laboratory}\\
       \affaddr{CCS-7, MS B287}\\
       \affaddr{Los Alamos, NM 87545}\\
       \email{crasmussen@lanl.gov}
% 3rd. author
\alignauthor
Wayne N. Weseloh\\
       \affaddr{Los Alamos National Laboratory}\\
       \affaddr{XTD-1, MS T082}\\
       \affaddr{Los Alamos, NM 87545}\\
       \email{weseloh@lanl.gov}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor
Robert W. Robey\\
       \affaddr{Los Alamos National Laboratory}\\
       \affaddr{XCP-2, MS T086}\\
       \affaddr{Los Alamos, NM 87545}\\
       \email{brobey@lanl.gov}
% 5th. author
\alignauthor
Daniel Quinlan\\
       \affaddr{Lawrence Livermore National Laboratory}\\
       \email{dquinlan@llnl.gov}
% 6th. author
\alignauthor
Jeffrey Overbey\\
       \affaddr{University of Illinois at Urbanna-Champaign}\\
       \email{overbey2@illinois.edu}
}

@@

\begin{abstract}
Emerging GPU architectures for high performance computing are well suited to a
data-parallel programming model.  This paper presents a simple programming
methodology that allows programmers to take advantage of these emerging systems.
We use array constructs in Fortran 90 to show how this infrequently exploited,
standardized language feature is easily transformed to lower-level accelerator
code.  These transformations are based on a simple mapping from Fortran 90 to
OpenCL.  We demonstrate, using a stencil code solving the shallow-water fluid
equations, that the performance of our compiler-generated transformations is
comparable with that of hand-optimized OpenCL code.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

>>>>

\title{ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming}

@@

\author{
% 1st. author
\alignauthor
Matthew J. Sottile\\
       \affaddr{Galois, Inc.}\\
       \affaddr{421 SW 6th Ave. Suite 300}\\
       \affaddr{Portland, OR 97204}\\
       \email{matt@galois.com}
% 2nd. author
\alignauthor
Craig E Rasmussen\\
       \affaddr{Los Alamos National Laboratory}\\
       \affaddr{CCS-7, MS B287}\\
       \affaddr{Los Alamos, NM 87545}\\
       \email{crasmussen@lanl.gov}
% 3rd. author
\alignauthor
Wayne N. Weseloh\\
       \affaddr{Los Alamos National Laboratory}\\
       %\affaddr{XTD-1, MS T082}\\
       %\affaddr{Los Alamos, NM 87545}\\
       \email{weseloh@lanl.gov}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor
Robert W. Robey\\
       \affaddr{Los Alamos National Laboratory}\\
       %\affaddr{XCP-2, MS T086}\\
       %\affaddr{Los Alamos, NM 87545}\\
       \email{brobey@lanl.gov}
% 5th. author
\alignauthor
Daniel Quinlan\\
       \affaddr{Lawrence Livermore National Laboratory}\\
       \email{dquinlan@llnl.gov}
% 6th. author
\alignauthor
Jeffrey Overbey\\
       \affaddr{University of Illinois at Urbanna-Champaign}\\
       \email{overbey2@illinois.edu}
}

@@

\begin{abstract}
  Emerging GPU architectures for high performance computing are well suited to a
  data-parallel programming model.  This paper presents preliminary work
  examining a programming methodology that provides Fortran programmers with access
  to these emerging systems.  We use array constructs in Fortran to
  show how this infrequently exploited, standardized language feature is easily
  transformed to lower-level accelerator code.  The transformations in ForOpenCL are based
  on a simple mapping from Fortran to OpenCL.  We demonstrate, using a
  stencil code solving the shallow-water fluid equations, that the performance
  of the ForOpenCL compiler-generated transformations is comparable with that of
  hand-optimized OpenCL code.
\end{abstract}

% A category with the (minimum) three required fields
\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This research was supported by the Departement of Energy, Office of Science.

### 2011-06-26 papers/GPUScA-2011/intro.tex
Version to ship to Matt.  Needs a couple of bib entries.

The peak performance of the newer accelerator architectures can be substantial.
Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
processor while the performance of the M2090 NVIDIA Tesla processor is in the
same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
many of the new accelerator architectures offer comes at a cost.
Architectural changes are trending toward multiple heterogeneous cores and less
of a reliance on superscalar instruction level parallelism and hardware managed
memory hierarchies (such as traditional caches).  These changes place a heavy
burden on application programmers as they are forced to adapt to these new
systems.  An especially challenging problem is not only how to program to these
new architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming models.  How
can a programmer easily write one program that can perform well on both a
conventional multicore CPU and a GPU (or one of the emerging many-core
architectures)?  A directive-based approach, such as OpenMP or the Accelerator
programming model from the Portland Group~\cite{pgi10accelerator}, is one
solution to this problem.

In this paper we take a somewhat different approach.  A common theme amongst the
new processors is the emphasis on data-parallel programming.  This model is well
suited to the emerging architectures that are based on either vector processing
or massively parallel collections of simple cores.  The recent CUDA and OpenCL
programming languages are are intended to support this programming model,

@@

We demonstrate that the data-parallel features of Fortran map surprisingly well
onto GPUs when transformed to OpenCL kernels.  These features include: array
syntax (e.g., assignment and binary operators), array constructs like {\tt
  WHERE} and {\tt DO CONCURRENT}, and the use of pure and elemental functions.
In addition, we provide new functions that explicitly take advantage of the
stencil geometry of the problem domain we are addressing.  Note that this subset
of the Fortran language is implicitly parallel.  This programming model
\emph{does not require explicit declaration of parallelism within the program.}
In addition, programs are expressed using entirely standard Fortran so it can be
compiled for and executed on a single core without concurrency.

@@

Transformations for arbitrary Fortran procedures are not attempted.
Furthermore, a mechanism to transform the calling site to automatically invoke
OpenCL kernels is not provided at this time.  While it is possible to accomplish
this task within ROSE, it is considered outside the scope of this paper.
However, we do provide (via Fortran interfaces) a mechanism to call the C OpenCL
runtime and enable Fortran programmers to access OpenCL kernels generated by the
automatic transformations.

>>>>

The peak performance of the newer accelerator architectures can be substantial.
Intel expects a teraflop for the SGEMM benchmark with their Knights Ferry
processor while the performance of the M2090 NVIDIA Tesla processor is in the
same neighborhood~\cite{hpcwire11manycore}.  Unfortunately the performance that
many of the new accelerator architectures offer comes at a cost.  Architectural
changes are trending toward multiple heterogeneous cores and less of a reliance
on superscalar instruction level parallelism and hardware managed memory
hierarchies (such as traditional caches).  These changes place a heavy burden on
application programmers as they are attempt to adapt to these new systems.  An
especially challenging problem is not only how to program to these new
architectures --- considering the massive scale of concurrency available --- but
also how to design programs that are portable across the changing landscape of
computer architectures.  How does a programmer write one program that can
perform well on both a conventional multicore CPU \emph{and} a GPU (or another
of the emerging many-core architectures)?

A directive-based approach, such as OpenMP or the Accelerator programming model
from the Portland Group~\cite{pgi10accelerator}, is one solution to this
problem.  However, in this paper we take a somewhat different approach.  A
common theme amongst the new processors is the emphasis on data-parallel
programming.  This model is well suited to the emerging architectures that are
based on either vector processing or massively parallel collections of simple
cores.  The recent CUDA and OpenCL programming languages are are intended to
support this programming model.

@@

We demonstrate that a subset of Fortran map surprisingly well onto GPUs when
transformed to OpenCL kernels.  This data-parallel subset includes: array syntax
using assignment statements and binary operators, array constructs like {\tt
  WHERE}, and the use of pure and elemental functions.  In addition, we provide
new functions that explicitly take advantage of the stencil geometry of the
problem domain we consider.  Note that this subset of the Fortran language is
implicitly parallel.  This programming model \emph{does not require explicit
  declaration of parallelism within the program.}  In addition, programs are
expressed using entirely standard Fortran so it can be compiled for and executed
on a single core without concurrency.

@@

Transformations for arbitrary Fortran procedures are not attempted.
Furthermore, a mechanism to transform the calling site to automatically invoke
OpenCL kernels is not provided at this time.  While it is possible to accomplish
this task within ROSE, it is considered outside the scope of this paper.
However, ForOpenCL provides via Fortran interfaces a mechanism to call the C OpenCL
runtime and enable Fortran programmers to access OpenCL kernels generated by the
supplied transformations.

### 2011-06-26 papers/GPUScA-2011/programming-model.tex
Version to ship to Matt.  Needs a couple of bib entries.

A question that one may pose is ``\emph{Why Fortran and not a more modern
  language like X?}''  The recent rise in interest in concurrency and
parallelism at the language level driven by multicore CPUs and manycore
accelerators has driven a number of new language developments, both as novel
languages and extensions on existing ones.  For many scientific users familiar
with Fortran, new languages and language extensions to use novel new
architectures present a challenge: how do they effectively use them while
avoiding rewriting code and potentially growing dependent on a transient
technology that will vanish tomorrow?

In this section we present the Fortran subset employed in this paper.  This
subsetting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We first examine how this
programming model in Fortran relates to developments in other languages.

@@

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures, particularly in previous decades
during the reign of vector and massively parallel computers in high performance
computing (HPC).  The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops (such
as within IVTRAN for the Illiac IV).

@@

Similar whole array shifting was used in higher dimensions for finite
difference codes within the computational physics community, especially
at Los Alamos for codes targeting the CM-5 system that resided there until the
late 1990s.  A body of research in compilation of stencil-based codes
that use shift operators targeting these systems is related to the
work we present here~\cite{stencil-compiler}.

The whole-array model was attractive because it deferred responsibility for
optimally implementing the computations to the compiler.  Instead of relying on
a compiler to infer parallelism from a set of explicit loops, the choice for how
to implement loops was left entirely up to the tool.  Unfortunately, this had
two side effects that have limited broad acceptance of the whole-array
programming model in Fortran.  First, programmers must translate their
algorithms into a set of global operations.  Finite difference stencils and
similar computations are traditionally defined in terms of offsets from some
central index.  Shifting, while conceptually analogous, can be awkward to think
about for high dimensional stencils with many points.  Second, the semantics of
these operations are such that all elements of an array operation are updated as
if they were updated simultaneously.  In a program where the programmer
explicitly manages arrays and loops, double buffering techniques and user
managed temporaries are used to maintain these semantics.  When the compiler is
responsible for managing this intermediate storage, it has historically proven
that they are inefficient and generate code that requires far more temporary
storage than really necessary.  This is not a flaw of the language constructs,
but a sign of the lack of sophistication of the compilers with respect to their
internal analysis to determine how to optimally generate this intermediate
storage.

An interesting line of language research that grew out of the early work with
HPF was that associated with the ZPL language work at the University of
Washington~\cite{chamberlain04zpl}.  In ZPL, programmers adopt a similar global
view of computation over arrays, but define their computations based on regions,
which provide a local view of the set of indices that participate in the update
of each element of an array.  The Chapel parallel programming language being
developed by Cray is directly influenced by HPF and
ZPL~\cite{chamberlain_chapel}.

@@

The static analysis and source-to-source transformations used in this work are
very basic and only require the programmer to use a language subset that employs
a data-parallel programming model.  In particular, it encourages the use of
array notation, elemental functions, and pure procedures.  From these language
constructs, we are able to easily transform Fortran procedures to a lower-level
OpenCL kernel implementation.

@@

Fortran has a rich array syntax that allows programmers to write statements in
terms of whole arrays or subarrays, with data parallel operators to compute on
the arrays.  Array variables can be used in expressions based on whole-array
operations.  For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays of the
same rank and shape and {\tt s} is a scalar, then the statement

@@

\noindent results in the element-wise sum of {\tt A} and {\tt s} times the
elements of {\tt B} being stored in the corresponding elements of {\tt C}. The
first element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

@@

An elemental function consumes and produces scalar values, but can be applied to
variables of array type such that the function is applied to each and every
element of the array.  This allows programmers to avoid explicit looping and
instead simply state that they intend a function to be applied to every element
of an array in parallel, deferring the choice of implementation technique to the
compiler.  Elemental functions are intended to be used for data parallel
programming, and as a result must be side effect free and mandate an {\tt
intent(in)} attribute for all arguments.

@@

Since pure procedures don't have side effects they are candidates for running on
accelerators in OpenCL.  Currently our prototype tool transforms pure procedures
to OpenCL kernels that \emph{do not} call other procedures, except for elemental
functions.

@@

Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
set of functions that allow programmers to work with subarrays in expressions.
In Fortran, these functions return a copy of an existing array or array section.
This is unlike ZPL, where regions are analogous to index sets and are used
primarily for address resolution within an array without dictating storage
related behavior.  The functions that we propose are similar in that they allow
a programmer to deal with index regions that are meaningful to their algorithm,
and automatically induce a halo (or ghost) cell pattern as needed in the
implementation generated by the compiler, where the size of an array is
increased to provide extra array elements surrounding the interior portion of
the array.  This is critical to reducing the amount of code that the programmer
is forced to write, as the halo cells are often where boundary errors and
off-by-one errors may occur in code that is manually generated.

Region functions are similar to the shift operator as they can be used to
reference portions of the array that are shifted with respect to the interior
portion.  However, unlike the shift operator, regions are not expressed in terms
of boundary conditions and thus don't explicitly \emph{require} a knowledge of
nor the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which access
only local subsections of an array stored in global memory.

ForOpenCL provides two new procedures that are defined in Fortran and are used
in array-syntax operations.  Each procedure takes a integer array halo argument
that specifies the number of ghost cells on either side of a region, for each
dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
for a two-dimensional region.  These functions are:

@@

\item {\tt interior(array, halo)}: a pure function that returns a copy of
  the interior portion of the array specified by halo.

\item {\tt region(array, halo)}: a pure function that returns a copy of
  the portion of the array specified by halo.

@@

It should be noted that the two functions {\tt interior} and {\tt region} are
pure and thus can be called from within a pure kernel procedure.  These two
functions are part of the language recognized by the compiler and though the two
functions return a copy of a portion of an array \emph{semantically}, the
compiler is not forced to actually make a copy and is free to enforce copy
semantics through other means.  In addition to these two new functions,
ForOpenCL provides the compiler directive, {\tt \!\$OFP KERNEL}, which specifies
that a pure subroutine can be transformed to an OpenCL kernel.

@@

There are several advantages to this style of programming using array
syntax, shifts, regions, and pure and elemental functions:

@@

Data parallelism has been called collection-oriented programming by
Blelloch~\cite{blelloch90}.  As the {\tt cshift} function and the array-valued
expressions all semantically return a value, this style of programming is also
similar to functional programming (or value-oriented programming).

Unfortunately, this style of programming has never really caught on because when
array syntax was first introduced in Fortran, performance of these features was
relatively poor and thus programmers shied away from using array syntax (even
recently, some are actively counseling against its usage because of performance
issues~\cite{Levesque:SC08}).  Thus the Fortran community was caught in a
classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it
was slow; and (2) compilers vendors didn't improve it because programmers didn't
use it.  A goal of this paper is to demonstrate that parallel programs written
in this style of Fortran are maintainable and can achieve good performance on
accelerator architectures.

@@

Only pure Fortran procedures are transformed into OpenCL kernels.  The
programmer must explicitly call these kernels from Fortran using the ForOpenCL
library described below.  It is also possible, using ROSE, to modify the calling
site so that the entire program can be transformed, but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively new
to ROSE, some generated code must be modified.  For example, the {\tt
  \_\_global} attribute for kernel arguments was added by hand.

@@

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel procedure
may not call other procedures except for limited intrinsic functions (primarily
math), user-defined elemental functions, and the {\tt interior} and {\tt region}
functions.  Future work will address non-contiguous arrays (such as those that result
from strided access) by mapping array strides to gather/scatter-style memory
accessors.

>>>>

A question that one may pose is ``\emph{Why choose Fortran and not a more modern
  language like X for programming accelerator architectures?}''  The recent rise
in interest in concurrency and parallelism at the language level driven by
multicore CPUs and manycore accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.  However,
for many scientific users with existing codes written in Fortran, new languages
and language extensions to use novel new architectures present a challenge: how
do they effectively use them while avoiding rewriting code and potentially
growing dependent on a transient technology that will vanish tomorrow?
In this paper we explore the constructs in Fortran that are particularly
relevant to GPU architecures.

In this section we present the Fortran subset employed in this paper.  This
subsetting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We start by examining how this
programming model relates to developments in other languages.

@@

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures.
%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops (e.g.,
IVTRAN for the Illiac IV).

@@

\noindent
Similar whole array shifting was used in higher dimensions for finite difference
codes within the computational physics community for codes targeting the CM-5
system.  Research in compilation of stencil-based codes that use shift operators
targeting these systems is related to the work presented
here~\cite{stencil-compiler}.

The whole-array model was attractive because it deferred responsibility for
optimally implementing the computations to the compiler.  Instead of relying on
a compiler to infer parallelism from a set of explicit loops, the choice for how
to implement loops was left entirely up to the tool.

Unfortunately, this had
two side effects that have limited broad acceptance of the whole-array
programming model in Fortran.  First, programmers must translate their
algorithms into a set of global operations.  Finite difference stencils and
similar computations are traditionally defined in terms of offsets from some
central index.  Shifting, while conceptually analogous, can be awkward to think
about for high dimensional stencils with many points.  Second, the semantics of
these operations are such that all elements of an array operation are updated as
if they were updated simultaneously.  In a program where the programmer
explicitly manages arrays and loops, double buffering techniques and user
managed temporaries are used to maintain these semantics.  When the compiler is
responsible for managing this intermediate storage, it has historically proven
that they are inefficient and generate code that requires far more temporary
storage than really necessary.  This is not a flaw of the language constructs,
but a sign of the lack of sophistication of the compilers with respect to their
internal analysis to determine how to optimally generate this intermediate
storage.

An interesting line of language research that grew out of HPF was that
associated with the ZPL language at the University of
Washington~\cite{chamberlain04zpl} and Chapel, an HPCS language developed by
Cray~\cite{chamberlain_chapel}.  In ZPL, programmers adopt a similar global view
of computation over arrays, but define their computations based on regions,
which provide a local view of the set of indices that participate in the update
of each element of an array.

@@

The static analysis and source-to-source transformations used in this work
require the programmer to use a language subset that employs a data-parallel
programming model.  In particular, it encourages the use of array notation,
elemental functions, and pure procedures.  From these language constructs, we
are able to easily transform Fortran procedures to a lower-level OpenCL kernel
implementation.

@@

Fortran has a rich array syntax that allows programmers to write statements in
terms of whole arrays or subarrays, with data-parallel operators to compute on
the arrays.  Array variables can be used in expressions based on whole-array
operations.  For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays of the
same rank and shape and {\tt s} is a scalar, then the statement

@@

\noindent
results in the element-wise sum of {\tt A} and the product of {\tt s} times the
elements of {\tt B} being stored in the corresponding elements of {\tt C}. The
first element of {\tt C} will contain the value of the first element of {\tt A}
added to the first element of {\tt c*B}.  Note that no explicit iteration over
array indices is needed and that the individual operators, plus, times, and
assignment are applied by the compiler to individual elements of the arrays
independently.  Thus the compiler is able to spread the computation in the
example across any hardware threads under its control.

@@

An elemental function consumes and produces scalar values, but can be applied to
variables of array type such that the function is applied to each and every
element of the array.  This allows programmers to avoid explicit looping and
instead simply state that they intend a function to be applied to every element
of an array in parallel, deferring the choice of implementation technique to the
compiler.  Elemental functions are intended to be used for data-parallel
programming, and as a result must be side effect free and mandate an {\tt
intent(in)} attribute for all arguments.

@@

Since pure procedures don't have side effects they are candidates for running on
accelerators in OpenCL.  Currently our prototype tool only transforms pure procedures
to OpenCL kernels that \emph{do not} call other procedures, except for elemental
functions, either defined by the user or intrinsic to Fortran.

@@

Borrowing ideas from ZPL, we introduce a concept of a region to Fortran with a
set of functions that allow programmers to work with subarrays in expressions.
In Fortran, these functions return a copy of or a pointer to an existing array or array section.
This is unlike ZPL, where regions are analogous to index sets and are used
primarily for address resolution within an array without dictating storage
related behavior.  The functions that we propose are similar in that they allow
a programmer to deal with index regions that are meaningful to their algorithm,
and automatically induce a halo (or ghost) cell pattern as needed in the
implementation generated by the compiler, where the size of an array is
explicitly increased to provide extra array elements surrounding the interior portion of
the array.
%%%This is critical to reducing the amount of code that the programmer is forced to write, as the halo cells are often where boundary errors and off-by-one errors may occur in code that is manually generated.

Region functions are similar to the shift operator as they can be used to
reference portions of the array that are shifted with respect to the interior
portion.  However, unlike the shift operator, regions are not expressed in terms
of boundary conditions and thus don't explicitly \emph{require} a knowledge of,
nor the application of, boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which access
only local subsections of an array stored in global memory.

ForOpenCL provides two new functions that are defined in Fortran and are used
in array-syntax operations.  Each procedure takes a integer array halo argument
that specifies the number of ghost cells on either side of a region, for each
dimension.  For example {\tt halo = [left, right, down, up]} specifies a halo
for a two-dimensional region.  These functions are:

@@

\item {\tt region\_cpy(array, halo)}: a pure function that returns a copy of
  the interior portion of the array specified by halo.

\item {\tt region\_ptr(array, halo)}: an impure function that returns a pointer 
  to the portion of the array specified by halo.

@@

It should be noted that the function {\tt region\_cpy} is pure and thus can be
called from within a pure kernel procedure, though it should be noted that {\tt
region\_ptr} is impure because it aliases the array parameter.  However as will
be shown below, the usage of {\tt region\_ptr} is constrained so that it does not
introduce side effects in the functions that call it.  These two functions are
part of the language recognized by the compiler and though {\tt region\_cpy}
returns a copy of a portion of an array \emph{semantically}, the compiler is not
forced to actually make a copy and is free to enforce copy semantics through
other means.  In addition to these two new functions, ForOpenCL provides the
compiler directive, {\tt \!\$OFP PURE, KERNEL}, which specifies that a pure subroutine
can be transformed to an OpenCL kernel and that the subroutine is pure
except for calls to {\tt region\_ptr}.

@@

There are several advantages to this style of programming using array
syntax, regions, and pure and elemental functions:

@@

Data parallelism has been called collection-oriented programming by
Blelloch~\cite{blelloch90}.  As the {\tt cshift} function and the array-valued
expressions all semantically return a value, this style of programming is also
similar to functional programming (or value-oriented programming).  It should be
noted that the subsetting language we employ goes beyond pure data parallelism
by the use of pure (other than calls to {\tt region\_ptr}) subroutines and
not just elemental functions.

Unfortunately, this style of programming has never really caught on because when
array syntax was first introduced in Fortran, performance of codes using these features was
relatively poor and thus programmers shied away from using array syntax (even
recently, some are actively counseling against its usage because of performance
issues~\cite{Levesque:SC08}).  Thus the Fortran community was caught in a
classic ``chicken-and-egg'' conundrum: (1) programmers didn't use it because it
was slow; and (2) compilers vendors didn't improve it because programmers didn't
use it.  A goal of this paper is to demonstrate that parallel programs written
in this style of Fortran can achieve good performance on accelerator architectures.

@@

Only pure Fortran procedures are transformed into OpenCL kernels.  This restriction
is lifted slightly to allow calls to {\tt region\_ptr} from within a kernel procedure.  The
programmer must explicitly call these kernels using Fortran interfaces in the ForOpenCL
library (described below).  It is also possible, using ROSE, to modify the calling
site so that the entire program can be transformed, but this functionality is
outside the scope of this paper.  Here we specifically examine transforming
Fortran procedures to OpenCL kernels.  Because OpenCL support is relatively new
to ROSE, some generated code must be modified.  For example, the {\tt
  \_\_global} attribute for kernel arguments was added by hand.

@@

Array variables within a kernel procedure (specified by the {\tt \!\$OFP PURE,
  KERNEL} directive) must be declared as contiguous.  A kernel procedure may not
call other procedures except for limited intrinsic functions (primarily math),
user-defined elemental functions, and the {\tt region\_cpy} and {\tt region\_ptr}
functions.  Future work will address non-contiguous arrays (such as those that
result from strided access) by mapping array strides to gather/scatter-style
memory accessors.

Array parameters to a kernel procedure must be declared as either intent(in) or
intent(out); they cannot be intent(inout).  A thread may read from an extended
region about its local element (using the {\tt region\_cpy} function, but can
only write to the single array element it owns.  If a variable were intent(inout),
a thread could update its array element before another thread had read from that
element.  This restriction requires double buffering techniques.

### 2011-06-26 papers/GPUScA-2011/shallow-water.tex
Version to ship to Matt.  Needs a couple of bib entries.

\noindent
where {\tt dx, dy, dt} are differential quantities in space $x, y$ and time $t$,
{\tt H, U}, and {\tt V} are state variables for the height and $x$ and $y$
momentum respectively, and {\tt oH, oU, oV} are corresponding output arrays used
in the double buffering scheme.  The \emph{OFP} compiler directive attributes
{\tt PURE} and {\tt KERNEL} indicate that the procedure {\tt wave\_advance} is
to be transformed as an OpenCL kernel and that it must be pure, other than for any
pointers used to reference interior regions of the output arrays.

>>>>

\noindent
where {\tt dx, dy, dt} are differential quantities in space $x, y$ and time $t$,
{\tt H, U}, and {\tt V} are state variables for the height and $x$ and $y$
momentum respectively, and {\tt oH, oU, oV} are corresponding output arrays used
in the double buffering scheme.  The \emph{OFP} compiler-directive attributes
{\tt PURE} and {\tt KERNEL} indicate that the procedure {\tt wave\_advance} is
to be transformed as an OpenCL kernel and that it must be pure, other than for any
pointers used to reference interior regions of the output arrays.
